---
title: "Lab 3: COVID-19"
subtitle: "Ecosystem Science and Sustianabilty 330"
author:
  - name: Genesis Lacy 
    email: glacy19@colostate.edu
output-dir: "docs"

format: 
  html:
    self-contained: true
    
execute: 
  echo: true
editor: visual
---

Q1: Easy access to open data is important to provide transparency with the public as well as providing data for those interested in the results and for people who are interested and/or are working on data analysis.

```{r}
url = "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
read.csv(url)


```

Q2: Creating a subset to generate cumulative cases and new cases for the top 5 counties.

```{r}
library(tidyverse); library(flextable)
data = read_csv('https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv')

my.date <- as.Date("2022-02-01")
my.state <- "Colorado"

co_data <- data %>%
  filter(state == my.state) %>%
  arrange(county, date) %>%
  group_by(county) %>%
  mutate(new_cases = cases - lag(cases, default = first (cases)),
        new_deaths = deaths - lag (deaths, default = first(deaths))) %>%
  ungroup()

```

```{r}
latest_data <- co_data %>%
  filter(date == my.date)

top5_cumulative_cases <- latest_data %>%
  arrange(desc(cases)) %>%
  slice_head(n = 5) %>%
  select(county, cases)
  
top5_new_cases <- latest_data %>%
  arrange(desc(new_cases)) %>%
  slice_head(n = 5) %>%
  select(county, new_cases)

top5_cumulative_cases %>%
  flextable() %>%
  set_caption("Top 5 counties by cumulative cases")

top5_new_cases %>%
  flextable() %>%
  set_caption("Top 5 counties by new cases")
```

Q3:Normalizing Data - adding population data


```{r}
pop_url <- 'https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/totals/co-est2023-alldata.csv'
pop_data <- read_csv(pop_url)

pop_data <- pop_data %>%
  mutate(
    STATE = str_pad(as.character(STATE), 2, pad = "0"),
    COUNTY = str_pad(as.character(COUNTY), 3, pad = "0"),
    FIPS = paste0(STATE, COUNTY)
  
  ) %>%
  filter(COUNTY != "000") %>%
  select(FIPS,STNAME,CTYNAME, contains("2021"))

glimpse(pop_data)



```

After Modification we're only looking at columns we filtered to contain "2021" and "name". A lot of the values are doubles and there are also a few characters. 

What are the range of populations seen in Colorado counties in 2021?

```{r}
co_data <- co_data %>%
  left_join(pop_data, by = c("fips" = "FIPS"))

co_data <- co_data %>%
  mutate(
    pop_2021 = coalesce(POPESTIMATE2021.x, 1) ,
    cumulative_per_capita = cases / pop_2021,
    new_cases_per_capita = new_cases / pop_2021, 
    new_deaths_per_capita = new_deaths / pop_2021
  )
  
  latest_data <- co_data %>%
    filter(date == my.date)
  
top_cumulative_per_capita <- latest_data%>%
  arrange(desc(cumulative_per_capita))%>%
  slice_head(n = 5)%>%
  select(county, cumulative_per_capita)%>%
  flextable()%>%
  set_caption("Top 5 Counties by Cumulative Cases per Capita")

top_new_per_capita <- latest_data%>%
  arrange(desc(new_cases_per_capita))%>%
  slice_head(n = 5)%>%
  select(county, new_cases_per_capita)%>%
  flextable()%>%
  set_caption("Top 5 Counties by New Cases per Capita")

top_cumulative_per_capita
top_new_per_capita
```

Q4:




